Statistical methods for relative quantification of post-translational modifications in global proteomics experiments

### 2022-01-28 main/MSstatsPTM_Main.tex
Changes for internal review

@@

The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD)\cite{LUCHETTI2021}. Multiplex proteomics was used to quantify the abundance of total protein and ubiquitination in human epithelial cells. Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.

@@

The proposed approach first leverages the large amount of work done to create MSstats~\cite{Choi:2014}. It takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}. This approach allows the method to be extended to cases where the experimental design in inbalanced, and where additional sources of variation are present. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}. After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models\cite{Bolker2009} \cite{Faraway:2006}. Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

@@

\begin{equation}
\begin{aligned}
H_{0}: \Delta = ( \mu_{i}^{\ast} - \mu_{i'}^{\ast} ) - ( \mu_{i} - \mu_{i'} ) = 0 \\
H_{a}: \Delta = ( \mu_{i}^{\ast} - \mu_{i'}^{\ast} ) - ( \mu_{i} - \mu_{i'} ) \neq 0
\end{aligned}
\end{equation}

@@

Model inference is done in the same way to how MSstats~\cite{Choi:2014} and MSstatsTMT~\cite{Huang:2020} target the global protein. In cases where the design in unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model. Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:adjustment_summary}.

@@

The proposed methods are implemented in the open source R package $MSstatsPTM$, available on Bioconductor. $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut. The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$. Conversion is done separately for the modified and global profiling runs. If the global profiling run is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance.

@@

The performance of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known. For the experiments where ground truth is known, we calculated summary statistics using the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN). To evaluate the performance of different methods, we considered the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and total accuracy ((TP + TN) / (TP + TN + FP + FN)). For biological experiments where ground truth was not known, we evaluated significantly differential peptides with $\alpha=.05$. Comparisons were made before and after adjusting for changes in unmodified protein abundance and the results are presented in the form of venn diagrams and profile plots. All datasets in this paper were evaluated using the $MSstatsPTM$ package.

@@

In this simulation, changes in unmodified peptide level still needed to be accounted for to control the FDR. Once controlling for changes in unmodified peptide, it is clear that the proposed method performed strongest across all methods, as seen in Figure \ref{fig:sim_accuracy}. The proposed method well calibrated model accuracy and the recall rate, even when the number of replicates were low. Finally, when comparing the fold change estimation across all peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change. In comparison, Limma and t-test showed much wider distributions. This fold change comparison can be seen in Figure \ref{fig:sim_boxplot}.

@@

The results of this experiment are summarized in Figures \ref{fig:ipah_venn_diagramm} and \ref{fig:ipah_profile}. The experiment was modeled as a group comparison. In Figure \ref{fig:ipah_venn_diagramm} the number of significant modified peptides before and after adjustment is shown. We can see that more peptides become insignificant after adjustment than become significant. 3,236 peptides become insignificant, 1352 peptides become significant, while 4,282 peptides are significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues are correct in adjustment, and the true abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was truly due to changes in global protein abundance.

In Figure \ref{fig:ipah_profile}, looking at modified peptide $GSDMD\_HUMAN|P57764$ we can see the advantage of the proposed method. The modified peptide shows a flat abundance change between the infected 1 hour, 4 hour, and 6 hour conditions. This is contrasted with a strong negative change in the global profiling run between the same conditions. Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$. After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became very significant, $5.25e^{-8}$. In this case the effect of the modified peptide is strongly confounded with changes in the global protein. The proposed method allows us to remove this confounding and estimate the true effect.

@@

The results of this experiment are summarized in Figures \ref{fig:shig_venn_d} and \ref{fig:shigella_profile}. The experiment was modeled as a time series, with the same bioreplicate measured at each time point. In Figure \ref{fig:shig_venn_d} the number of significant modified peptides before and after adjustment is shown. Again more PTMs became insignificant after adjustment than became significant. 19,286 peptides become insignificant, 4,947 became significant, while 41,552 peptides are significant in both models. Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment, or if it was mainly due to removing convolution with change in the umodified protein. This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance. This is still a very small portion (3.4\%) of the peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.65]{images/fig3.png}
\caption{
Data structure of a typical PTM experiment and goals of PTM characterization. (a) Schematic data representation, in a simplified case of two conditions and two replicate runs. Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide. The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i. Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by i*. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). Some spectral features can be missing. (b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates. The advantage of using the proposed approach is apparent when looking at the second computer simulation, which includes limited observations and the presence of missing values. In overall accuracy, the proposed method performs stronger than Limma and t-test in nearly every model. Even at lower replicates the proposed method still outperformed Limma. The lowest performaning method was $t$-test. Limma shows comparable performance to the proposed method in a clean experiment, however when real world data problems are introduced it is clear the proposed method is more robust.}
\label{fig:sim_accuracy}
\end{figure}

>>>>

There are two main biological questions that are generally relevant in PTM analysis. First, some researchers run mass-spectrometry based proteomic experiments targeting PTMs when looking for biomarkers\cite{Petushkova_2017}. In this case the researcher may not necessarily want to correct for confounding with the total protein. The interest is in detecting modified peptides which change in abundance in a meaningful way, such as between disease and control, fast or slow-progressing, or responder and non-responder groups. Conversely, researchers may be interested in whether the fraction of peptides modified at a given site is changing between conditions. This is an important question for thinking about biological and physiological models and very short timescale experiments. In this case, correcting for confounding with the total protein is relevant. These two questions are both important areas of study, and require different statistical methods to correctly address. In this paper we will mainly address a statistical solution to the second question, which requires more rigorous methods as we must address the convolution with the total protein.

@@

The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to degrade the protein gasdermin D (GSDMD)\cite{LUCHETTI2021}. Multiplex proteomics was used to quantify the abundance of total protein and ubiquitination in human epithelial cells. Cells were engineered to express IpaH7.8 under a dox inducible promoter and measurements were taken at different time periods. GSDMD was actively degraded when IpaH7.8 expression was induced by dox treatment. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.

@@

The proposed approach first leverages the large amount of work done to create MSstats~\cite{Choi:2014}. It takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}. This approach allows the method to be extended to cases where the experimental design is unbalanced, and where additional sources of variation are present. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}. After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models\cite{Bolker2009} \cite{Faraway:2006}. Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

@@

\begin{equation}
\begin{aligned}
H_{0}: \Delta = ( \mu_{i}^{\ast} - \mu_{i'}^{\ast} ) - ( \mu_{i} - \mu_{i'} ) = 0 \\
H_{a}: \Delta = ( \mu_{i}^{\ast} - \mu_{i'}^{\ast} ) - ( \mu_{i} - \mu_{i'} ) \neq 0
\end{aligned}
\label{eq_1}
\end{equation}

@@

Model inference is done in the same way to how MSstats~\cite{Choi:2014} and MSstatsTMT~\cite{Huang:2020} target the global protein. In cases where the design is unbalanced, a linear mixed effects model is fit, which takes into account all potential sources of variation, and restricted maximum likelihood (RML) is used to estimate the parameters of each model. Once the parameters are estimated, we can combine the models using modified versions of the formulas in Section \ref{sec:adjustment_summary}.

@@

The proposed methods are implemented in the open source R package $MSstatsPTM$, available on Bioconductor. $MSstatsPTM$ includes converters for multiple spectral processing tools, including MaxQuant, Progenesis, and Spectronaut. The converters take as input the raw data from the tool, identify the modification site for modified peptide, and put the data into the correct format for analysis in $MSstatsPTM$. Conversion is done separately for the modified and global profiling runs. If the global profiling run is not available, the package can still analyze the modified run, but will do so without adjusting for changes in unmodified protein abundance. Specifically, if a global profiling run is not available the null and alternative hypothesis in Equation \ref{eq_1} will reduce to:

\begin{equation}
\begin{aligned}
H_{0}: \Delta = \mu_{i}^{\ast} - \mu_{i} = 0 \\
H_{a}: \Delta = \mu_{i}^{\ast} - \mu_{i} \neq 0
\end{aligned}
\label{eq_9}
\end{equation}

Which amounts to running a the process without accounting for protein adjustment, as seen in MSstats and MSstatsTMT \cite{Choi:2014} \cite{Huang:2020}.

@@

The performance of the proposed method was evaluated on simulated and spike in datasets with known ground truth, as well as biological experimental data where the ground truth was not known. For the experiments where ground truth is known, we calculated summary statistics using the number of true positives (TP), false positives (FP), true negatives (TN), and false negatives (FN). To evaluate the performance of different methods, we considered the false discovery rate (FDR = FP/(TP+FP)), the recall rate (TP/(TP + FN), and accuracy ((TP + TN) / (TP + TN + FP + FN)). For biological experiments where ground truth was not known, we evaluated significantly differential peptides with $\alpha=.05$. Comparisons were made before and after adjusting for changes in unmodified protein abundance and the results are presented in the form of venn diagrams and profile plots. All datasets in this paper were evaluated using the $MSstatsPTM$ package.

@@

In this simulation, changes in unmodified peptide level still needed to be accounted for to control the FDR. Once controlling for changes in unmodified peptide, it is clear that the proposed method performed strongest across all methods, as seen in Figure \ref{fig:sim_accuracy}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. Finally, when comparing the fold change estimation across all peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change. In comparison, Limma and t-test showed wider distributions. This fold change comparison can be seen in Figure \ref{fig:sim_boxplot}.

@@

The results of this experiment are summarized in Figures \ref{fig:ipah_venn_diagramm} and \ref{fig:ipah_profile}. The experiment was modeled as a group comparison. In Figure \ref{fig:ipah_venn_diagramm} the number of significant modified peptides before and after adjustment is shown. We can see that more peptides become insignificant after adjustment than become significant. 3,236 peptides become insignificant, 1352 peptides become significant, while 4,282 peptides are significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues are correct in adjustment, and the true abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was truly due to changes in global protein abundance.

In Figure \ref{fig:ipah_profile}, looking at modified peptide $GSDMD\_HUMAN|P57764$ we can see the advantage of the proposed method. The modified peptide shows a flat abundance change between the infected 1 hour, 4 hour, and 6 hour conditions. This is contrasted with a strong negative change in the global profiling run between the same conditions. Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$. After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became very significant, $5.25e-8$. In this case the effect of the modified peptide is strongly confounded with changes in the global protein. The proposed method allows us to remove this confounding and estimate the true effect.

@@

The results of this experiment are summarized in Figures \ref{fig:shig_venn_d} and \ref{fig:shigella_profile}. The experiment was modeled as a time series, with the same bioreplicate measured at each time point. In Figure \ref{fig:shig_venn_d} the number of significant modified peptides before and after adjustment is shown. Again more PTMs became insignificant after adjustment than became significant. 19,286 peptides become insignificant, 4,947 became significant, while 41,552 peptides are significant in both models. Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment, or if it was mainly due to removing convolution with change in the umodified protein. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance. This is still a very small portion (3.4\%) of the peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.65]{images/fig3.png}
\caption{
Data structure of a typical PTM experiment and goals of PTM characterization. (a) Schematic data representation, in a simplified case of two conditions and two replicate runs. Each PTM site is modeled and characterized separately, where a PTM is quantified with multiple spectral features (boxes), distinguished by different charge states of a peptide. The feature intensities are viewed as repeated measurements of the underlying abundance of the PTM, where the abundance in Condition i is denoted by i*. Features corresponding to unmodified peptides are considered together to perform adjustment with respect to protein abundance, where the protein abundance in Condition i is denoted by i. Peptides can be fully cleaved (solid lines) and/or partially cleaved (dashed lines). Some spectral features can be missing. (b) PTM relative quantification by statistical inference, which makes use of the feature intensities to infer the underlying PTM abundance and protein abundance with an estimate of associated uncertainty. (c) Model-based testing for differential PTM abundance, which corrects for the underlying protein abundance with a cost of increased uncertainty about the estimate of difference between conditions.}
\label{fig:data-structure}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates. The advantage of using the proposed approach is apparent when looking at the second computer simulation, which includes limited observations and the presence of missing values. In overall accuracy, the proposed method performs stronger than Limma and t-test in nearly every model. Even at lower replicates the proposed method still outperformed Limma. The lowest performing method was $t$-test. Limma shows comparable performance to the proposed method in a clean experiment, however when real world data problems are introduced it is clear the proposed method is more robust.}
\label{fig:sim_accuracy}
\end{figure}

### 2022-01-28 supplementary/ptm_sm.tex
Changes for internal review

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim_new/spike_in_fc}
\caption{The fold change misalignment before adjustment is illustrated in the boxplots. The true log fold change is indicated by the red 'X' marks. Before adjustment the boxplots are not close to the true fold change. After adjustment all three models generally have a median near the true fold change. In terms of model comparison, the proposed method shows a much tighter distribution around the true fold change, whole both Limma and $t$-test are much wider. \label{fig:spike_fc}}
\end{figure}

@@

In this experiment Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD). \cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells. Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was inbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. The experiment was modeled as a group comparison, with different replicates measured at each time point. The experimental design can be seen in Table~\ref{table:ipah_design}.

@@

\begin{figure}[h!]
\centering
\includegraphics[height=.65\textwidth]{sim_new/shig_venn_diagramm.png}
\caption{The overlap of differential modified peptides between the PTM model with and without global protein level adjustment. Much like in \sfigref{fig:ipha_venn_d}, more PTMs became insignificant after adjustment then became significant. This indicates that for the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment. This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance. This is higher than in \sfigref{fig:ipha_venn_d}, however it is still a very small portion (3.4\%) of the peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.}
\label{fig:shig_venn_d}
\end{figure}

>>>>

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim_new/spike_in_fc}
\caption{The fold change misalignment before adjustment is illustrated in the boxplots. The true log fold change is indicated by the red 'X' marks. Before adjustment the boxplots are not close to the true fold change. After adjustment all three models generally have a median near the true fold change. In terms of model comparison, the proposed method shows a much tighter distribution around the true fold change, whole both Limma and $t$-test are wider. \label{fig:spike_fc}}
\end{figure}

@@

In this experiment Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD). \cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells. Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. The experiment was modeled as a group comparison, with different replicates measured at each time point. The experimental design can be seen in Table~\ref{table:ipah_design}.

@@

\begin{figure}[h!]
\centering
\includegraphics[height=.65\textwidth]{sim_new/shig_venn_diagramm.png}
\caption{The overlap of differential modified peptides between the PTM model with and without global protein level adjustment. Much like in \sfigref{fig:ipha_venn_d}, more PTMs became insignificant after adjustment then became significant. This indicates that for the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance. This is higher than in \sfigref{fig:ipha_venn_d}, however it is still a very small portion (3.4\%) of the peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.}
\label{fig:shig_venn_d}
\end{figure}
